---
title: "ICE Detention Guaranteed Minimums & Ghost Beds as of April 2021"
description: |
  An exploration of detention facility contract guaranteed minimums, where are they and what is the expected cost.
author:
  - name: Nathan Craig
    url: https://ncraig.netlify.app/
    affiliation: NMSU Anthropology, AVID
date: 04-28-2021
output:
  distill::distill_article:
    self_contained: false
    pandoc_args: ["--number-sections"]
    df_print: paged
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: yes
---

# Introduction {#introduction}

This post is an exploratory data analysis [@tukey1977] of ICE detention Guaranteed Minimums based on information reported in ICE's [FY21 year to date](https://www.ice.gov/detain/detention-management) report as of April 15, 2021 [@ice2021]. The data on this page are now updated regularly and may be subject to change in the future. This post examines one snapshot in time. Future posts may attempt to further track changes over time.

In addition to looking at Guaranteed Minimums, effort is made here to examine the number of empty Guaranteed Minimum beds which, following Selman and Leighton's [-@selman2010, 114] term of "ghost inmates" which originates in the private and for-profit corrections incarceration sector, are referred to here as "Ghost Beds." As Selman and Leighton note because private prisons operate and make money on economies of scale, many for-profit private prison contracts contain clauses that facilities must remain 90% full and if they aren't the contracts are structured to guaranteed payment for a minimum number of incarcerated persons. ICE facilities, both publicly and privately run, contain Guaranteed Minimums which are listed in the [FY21 year to date](https://www.ice.gov/detain/detention-management) report.

ICE's FY21 year to date report lists each facility's type, Average Daily Population (ADP) and contractual Guaranteed Minimums. Based on ADP and Guaranteed Minimums, it is possible to estimate the number of empty guaranteed beds or Ghost Beds. It is also possible to explore patterns of ADP, Guaranteed Minimums, and Ghost Beds by facility type.

## Summary {#summary}

For those who may not wish to read through the entire data exploration. A brief summary of some key findings are reported below.

Section 2 data exploration begins by looking at facility types, which appears in ICE's data set as Type Detailed, as well as ADP by Type Detailed. Private facilities are the most common facility type, though arrangements with US Marshals Service (USMS) are also extremely common. It bears mention that USMS contracts with ICE are under studied and extremely problematic [@jameson2020]. Not technically private contracts, they operate on a distinct business model related to procurement. While Service Processing Centers (SPC) have the highest ADP by type, Inter-Governmental Service Agreements (IGSA) and Contract Detention Facilities (CDF) are more numerous.

Section 3 looks at Guaranteed Minimums generally and by Type Detailed. At present, there are 34390 guaranteed beds distributed among 50 facilities. There is a strong correlation between total ADP, termed Total Detained and Guaranteed Minimum, Total Detained population explains 46% of the variation in Guaranteed Minimums. As a type, IGSA's have the largest share of Guaranteed Minimums overall, which is just under 2/3. On average, among facilities with Guaranteed Minimums, IGSA's tend to have larger Guaranteed Minimums and IGSA's are more numerous as a type. However, CDF's, with just under 1/3, also have a large portion of Guaranteed Minimums. IGSA's and CDF's make up 90% of the facilities with Guaranteed Minimums and 92% of the Guaranteed Minimum beds.

Section 4 looks at Ghost Beds generally, by Type Detailed and explores Guaranteed Minimum as well as Ghost Bed costs. The mean Guaranteed Minimum per diem is \$123.04 (SD=43.02). The total estimated cost of Guaranteed Minimums is \$4,308,391 per day, \$129,251,730 per month, or \$1,572,562,715 per year. At present 2/3 of Guaranteed Minimums are empty Ghost Beds. At the present rate of Ghost Beds, which may change, the total estimated cost is \$2,952,525 per day, \$88,575,750 per month, or \$1,077,671,625 per year.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Reading files
library(sf)
library(readr)
library(readxl)
library(rio)

# Data wrangling
library(tidyverse)
library(janitor)

# Graphing
library(ggplot2)
library(ggstatsplot)
library(ggforce)
library(ggthemes)

# Stats
library(easystats)
library(summarytools)

# Output
library(rmarkdown)
library(kableExtra)
library(knitr)
# knitr option that forces rounding
options(digits=3)
opts_chunk$set(results = 'asis',
               comment = NA,
               prompt = FALSE,
               cache = FALSE)
# knitr hook to put commas in the thousands place
# for inline numbers based on variables.
knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",")
})

# Turn off scientific notation
options(digits=5, scipen=15) 
```

```{r load-data, include=FALSE}
states <- st_read("C:/Users/nmc/Nextcloud/AVID_Restricted/ICE_data/ICE-Detention-Year-End-Reports/map_data/cb_2018_us_state_500k.shp")

df_attributes <-rio::import('https://github.com/n8craig/ICE-Detention-Year-End-Reports/raw/main/data/FY21_detention-stats_0414.xlsx', 
    sheet = "Facilities FY21 YTD", col_types = c("text", 
        "text", "text", "text", "numeric", 
        "text", "text", "text", "text", "text", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "text", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "text", "text", "text", 
        "date", "text", "text", "text", "date"), 
    skip = 6)

df_geocoded <- read_csv("https://raw.githubusercontent.com/n8craig/ICE-Detention-Year-End-Reports/main/data/facilities%20geocoded.csv")

df_geocoded <- df_geocoded %>% 
  select(1, 6:10)

df <- left_join(df_attributes, df_geocoded) %>% 
  relocate(32:36, .after=Zip) %>% 
  janitor::clean_names()

remove(df_attributes, df_geocoded)

# Set numeric fields
df$fy21_alos <- as.numeric(df$fy21_alos)
df$level_a <- as.numeric(df$level_a)
df$female_non_crim <- as.numeric(df$female_non_crim)

# Set factor fields
df$aor <- as.factor(df$aor)
df$type_detailed <- as.factor(df$type_detailed)
df$male_female <- as.factor(df$male_female)

# Create new columns
df <- df %>% 
  mutate_at(vars(14:27), funs(round(.,0))) %>% 
  mutate(total_detained=
           female_crim +
           female_non_crim +
           male_crim +
           male_non_crim) %>% 
  mutate(total_males_detained =
           male_crim +
           male_non_crim) %>% 
  mutate(total_females_detained =
           female_crim +
           female_non_crim) %>% 
  mutate(percent_no_threat =
            (no_ice_threat_level /
           total_detained)*100) %>% 
  mutate_at(vars(percent_no_threat), funs(round(.,0))) %>%  
  relocate(c(total_females_detained,
             total_males_detained,
             total_detained,
             percent_no_threat), .after = guaranteed_minimum)
```

# Facility Types {#sec:facility-types}

The [FY21 year to date](https://www.ice.gov/detain/detention-management) report lists 142 facilities, five of which are not located on the continental US. ICE lists six kinds of facilities:

-   Bureau of Prisons (BOP) are public facilities.
-   Contract Detention Facilities (CDF), these are private facilities generally owned by a private prison company.
-   Intergovernmental Service Agreement (IGSA), these are privately managed facilities owned by a county or municipality whose management is subcontracted to a private prison company.
-   Service Processing Centers (SPC), are public facilities managed by ICE though often the security staff are subcontracted to a private company.
-   US Marshals Service Intergovernmental Agreement (USMS IGA) are established US Marshals contracts with local jails that are used by ICE for immigration detention purposes. USMS IGA's may be a mix of public and private facilities depending on whether local governments subcontract out jail management to a private contractor.
-   Other

<aside>

The five ICE detention facilities located outside of the continental US are:

-   Honolulu Federal Detention Center, HI
-   Hiland Mountain Correctional Center, AK
-   Guaynabo MDC PR
-   Department Of Corrections GU
-   Saipan Department Of Corrections MP

</aside>

## Facility Counts by Facility Type {#facility-counts-by-facility-type}

Table \@ref(tab:facility-types) and Figure \@ref(fig:facility-types-bar-plot) show that by count the most common type of facility are IGSA's (n=73) followed by USMS IGAs (n=39), and CDF (n=17). IGSA's andCDF's are private and USMS IGA's may be a mix of public and private local jails (see list above). Regardless, between IGSA's and CDF's 63% (n=90) of ICE detention facilities by count are unambiguously privately managed. USMS IGA's still operate more as "procurement not policy" in what is an "opaque and intricate business practice" [@jameson2020: 283] in which local counties and municipalities view detention as a key source of income [@jameson2020: 2839].

```{r map-all-facilities, fig.cap="Map of ICE detention facilities."}
# Map of all facilities
ggplot()+
  geom_sf(data=states)+
  geom_point(data=df,
             aes(x=longitude, y=latitude), color = "Red", size = 1)+
    coord_sf(xlim = c(-165,150), ylim = c(10,72), expand = FALSE)+
  labs(
    title = "ICE Detention Facilities",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
    )
```

```{r facility-types}
# Facility count and percent total by facility type.
df_types <- df %>% 
  count(type_detailed) %>%
  mutate(percent_total_facilities = prop.table(n)*100) %>% 
  mutate(percent_total_facilities = round(percent_total_facilities,2))
# Sort by percent total
df_types %>% 
  arrange(percent_total_facilities)%>%
  kable(
    col.names = c("Type Detailed", "n",  "% Total Facilities"),
    caption="Count and Percent Total Facilities by Type."
    )
```

```{r facility-types-bar-plot, fig.cap="Bar chart of ICE detention facilities by facility type."}
# Bar Chart of Facility Count by Type
ggplot(df, aes(y = fct_infreq(type_detailed)))+
  geom_bar(stat = 'count')+
  ylab("Facility Type Detailed")+
  xlab("Count")+
  labs(
    title = "ICE Detention Center Facility Count by Facility Type",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
  )
```

Figure \@ref(fig:facility-map-facet-by-type) shows were ICE detention facilities are located based on facility type. IGSA's and USMS IGA's are more spread out while CDF's, SPC's, and Others are primarily located along the southern border.

```{r facility-map-facet-by-type, fig.cap="Facet map of ICE detention facilities by facility type."}
# Facet map of Facilities by Type
ggplot()+
  geom_sf(data=states)+
  geom_point(data=df,
             aes(x=longitude, y=latitude,
                 color = type_detailed), size = 2)+
  scale_shape_discrete(solid = FALSE)+
  coord_sf(xlim = c(-125,-64), ylim = c(16.5, 49.5), expand = TRUE)+
  scale_x_continuous(breaks = c(-120, -100, -80))+
  labs(
    title = "ICE Detention Facility Type",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
    )+
  facet_wrap(~type_detailed, ncol =3)
```

## ADP by Facility Type {#adp-by-facility-type}

```{r}
sum_adp <- sum(df$total_detained)
```

Based on ICE's FY2021 year to date data table, the total ADP is presently `r sum_adp`, which is generally consistent with the detained population reported on the ICE [Guidance on COVID-19](https://www.ice.gov/coronavirus) page for this time period. Examining total ADP by facility type shows that IGSA's (n=8999) and CDF's (n=3567), which are both private kinds of facilities, make up 81% (n=12566) of the total ADP (see Table \@ref(tab:adp-by-facility-type-table). However, IGSA's (n=73) and CDF's (n=17) make up 63% of the total number of facilities. They hold a larger portion of the total ADP because these kinds of facilities are generally larger, though at present SPC's appear to have a higher mean ADP though there are few of this kind of facility (n=5).

```{r adp-by-facility-type-table}
df_adp_summaries <- df %>%
group_by(type_detailed) %>% 
  summarize(total_adp = sum(total_detained),
            mean_adp = mean(total_detained),
            sd_adp = sd(total_detained)) %>% 
  mutate(percent_total_adp = (total_adp/sum_adp)*100) %>%
  mutate(percent_total_adp = round(percent_total_adp, digits=2)) %>% 
  mutate(mean_adp = round(mean_adp, digits = 2)) %>% 
  mutate(sd_adp = round(sd_adp, digits = 2)) %>%
  arrange(total_adp)

left_join(df_types, df_adp_summaries) %>% 
  kable(
    col.names = c("Type Detailed", "n", "% Total Facilities", "Total ADP", "Mean ADP", "SD ADP", "% Total ADP"),
    caption = "Summary statistics of ADP by ICE detention facility type."
    )
```

Looking at the distribution of ADP by ICE detention facility type, one can see that SPC's have the highest mean ADP (M=269.80, SD=121.26), while CDF's (M=209.82, SD=188.57) and IGSA's (M=123.27, SD=139.88) have the next highest mean ADP (see Table \@ref(tab:adp-by-facility-type-table)). A combined violin-sina plot (Figure \@ref(fig:adp-by-facility-type-figure)) illustrates the distribution of ADP by ICE detention facility type [@hintze1998; @sidiropoulos2018]. So while SPC's presently have larger detained populations on average, IGSA's and CDF's are far more numerous. While USMS IGA's (n= 39) are the second most numerous kind of facility they have a comparatively lower total ADP (n=1423) or by facility average ADP (M=36.49, SD=45.43).

```{r adp-by-facility-type-figure, fig.cap="Violin and Sina plot of ADPs by ICE detention facility type."}
df_mean <- df %>% 
  group_by(type_detailed) %>% 
  summarize(average = mean(total_detained)) %>% 
  ungroup()

df %>% 
  ggplot(data = ., aes(x= type_detailed, y = total_detained, fill=type_detailed))+
  geom_violin()+
  stat_summary(fun = mean, geom = "point", shape = 20, size = 5, color = "red", fill = "red")+
  geom_sina(alpha = 0.5)+
  geom_line(data = df_mean, mapping = aes(x = type_detailed, y = average, group = 1))+
  theme(legend.position = "none")+
  xlab("Type Detailed")+
  ylab("ADP")+
    labs(
    title = "Violin-Sina plot of Total ADP by ICE Detention Facility Type",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
    )
```

# Guaranteed Minimums {#guaranteed-minimums}

```{r guaranteed-min-var, include=FALSE}
# Variable of sum of guaranteed minimum beds, for use in text.
guaranteed_min_sum <- sum(df$guaranteed_minimum, na.rm = TRUE)
```

```{r include=FALSE}
cor.test(df$guaranteed_minimum, df$total_detained, method = "pearson") %>% 
  report()
```

```{r include=FALSE}
lm(guaranteed_minimum~total_detained, df) %>% 
  report()
```

Of the 142 facilities reported by ICE in the FY21 year to date dataset, 35% (n=50) have contractual guaranteed minimums (Table \@ref(tab:guaranteed-min-by-facility)). The Pearson's product-moment correlation between Guaranteed Minimum and Total Detained is positive, statistically significant, and very large ($r = 0.68, t(48) = 6.40, p < .001$). A linear model (estimated using OLS) was applied to predict Guaranteed Minimum with Total Detained. The model explains a statistically significant and substantial proportion of variance ($R^2 = 0.46, F(1, 48) = 40.91, p < .001, adj. R^2 = 0.45$). The effect of Total Detained on Guaranteed Minimums is statistically significant and positive ($b=2.07, t(48) = 6.40, p < .001$). Larger facilities have more guaranteed minimums.

## Guaranteed Minimums by Facility Type {#guaranteed-minimums-by-facility-type}

Facilities with guaranteed minimums are of three types: IGSA's, CDF's, and SPC's (Figure \@ref(fig:histogram-guaranteed-min-by-type) and Table \@ref(tab:guaranteed-min-by-facility-type)).

```{r guaranteed-min-by-facility, layout = "l-body-outset"}
# Guaranteed Minimums by Facility
df_guaranteed_min <- 
df %>% 
  as_tibble() %>% 
  select(c(name, type_detailed, guaranteed_minimum,
           total_detained)) %>% 
  filter(guaranteed_minimum >0) %>% 
  arrange(desc(guaranteed_minimum))

df_guaranteed_min %>% 
  kable(
    col.names = c("Facility", "Type Detailed", "Guaranteed\nMinimum", "Total Detained"),
    caption = "Guaranteed minimum and total detained by facility."
    ) %>% 
  scroll_box(height = "300px")
```

```{r histogram-guaranteed-min-by-type, fig.cap="Histogram of guaranteed minimums by ICE detention facility type."}

# Plot of Guaranteed Minimums
df_guaranteed_min %>% 
  ggplot()+
  geom_histogram(aes(x=guaranteed_minimum, color=type_detailed, fill = type_detailed),alpha=.6)+
  labs(title="Histogram of guaranteed minimum contracts showing the count \nof ICE detention facilities by type per number of guaranteed beds.")+
  ylab("Number of ICE detention facilities")
```

All 5 SPC's have guaranteed minimums, making up 10% of the facilities with guaranteed minimums, and forming 7.68% of all guaranteed minimum beds. SPC's average 528 guaranteed beds per facility which is the lowest of the three types. So SPC's have guaranteed minimums at the highest rate, but they make up a relatively small proportion of the facilities with guaranteed minimums, and a small proportion of the total number of guaranteed beds.

There are 30 IGSA's that have guaranteed minimums, 41% of this facility type has guaranteed minimums, they constitute 60% of the facilities with guaranteed minimums, which comprises 61.95 of all guaranteed minimum beds. IGSA's average 710.17 guaranteed beds per facility which is the highest of any of the three types. IGSA's constitute the majority of facilities with guaranteed minimums and the majority of guaranteed minimum beds. IGSA's also have the highest per facility guaranteed minimum average. The four facilities with the most number of guaranteed minimum beds are all IGSA's (Table \@ref(tab:guaranteed-min-by-facility).

There are 15 CFS that have guaranteed minimums, 88% of this type of facility has guaranteed minimums, they make up 10% of all facilities with guaranteed minimums , which comprises 30% of all guaranteed minimum beds, and as a type averages 696.2 guaranteed beds per facility. Thus, a very high proportion of this facility type has guaranteed minimums, but they make up a relatively small proportion of the total number of facilities with guaranteed minimums. CFS's have average guaranteed minimums that are relatively close to IGSA's.

Together, CDF's and IGSA's constitute 90% of the facilities with guaranteed minimums and 92% of the guaranteed minimum beds (Figure \@ref(fig:guaranteed-min-by-facility-type-figure)). The overwhelming number of guaranteed minimums are thus in private facilities.

```{r guaranteed-min-by-facility-type, layout = "l-body-outset"}

# Filter facility types to just those with guaranteed mins.
df_types_guaranteed_min <- 
df_types %>% 
  filter(type_detailed == "CDF" | type_detailed == "SPC" | type_detailed == "IGSA")

# Get counts and percent totals of facilities with guaranteed mins.
df_guaranteed_min_count <- df %>% 
  filter(guaranteed_minimum > 0) %>% 
  count(type_detailed) %>% 
  rename(n_facilities_w_guaranteed_min = n) %>% 
  mutate(percent_facilities_by_type_w_guaranteed_min = (n_facilities_w_guaranteed_min/df_types_guaranteed_min$n)*100) %>% 
  mutate(percent_facilities_by_type_w_guaranteed_min = round(percent_facilities_by_type_w_guaranteed_min, digits = 0)) %>% 
  mutate(percent_facilities_w_guaranteed_min = (n_facilities_w_guaranteed_min/50)*100)


# Get summary stats of guaranteed mins by facility type.
df_guaranteed_min_summaries <- df %>% 
  filter(guaranteed_minimum > 0) %>% 
  group_by(type_detailed) %>% 
  summarize(total_guaranteed_min = sum(guaranteed_minimum),
            mean_guaranteed_min = mean(guaranteed_minimum),
            sd_guaranteed_min= sd(guaranteed_minimum)) %>% 
   mutate(percent_total_guaranteed_min = (total_guaranteed_min/guaranteed_min_sum)*100, .after = total_guaranteed_min) %>% 
  mutate(percent_total_guaranteed_min = round(percent_total_guaranteed_min, digits = 2))

# Join the three tables
left_join(df_guaranteed_min_count, df_guaranteed_min_summaries) %>% 
  kable(
    col.names = c("Type Detailed",
                  "n\nFacilities w. Guaranteed\nMinimum",
                  "% Facilities by Type w. Guaranteed Minimum",
                  "% Facilities w. Guaranteed Minimum",
                  "Total Guaranteed Minimum",
                  "% Total Guaranteed Minimum",
                  "Mean Guaranteed Minimum",
                  "SD Guaranteed Minimum"),
    caption = "Guaranteed minimums by facility type."
    )
```

```{r guaranteed-min-by-facility-type-figure, fig.cap="Violin and Sina plot of guaranteed minimum by ICE detention facility type."}
df_mean <- df_guaranteed_min %>% 
  group_by(type_detailed) %>% 
  summarize(average = mean(guaranteed_minimum)) %>% 
  ungroup()

df_guaranteed_min %>% 
  ggplot(data = ., aes(x= type_detailed, y = guaranteed_minimum, fill=type_detailed))+
  geom_violin()+
  stat_summary(fun = mean, geom = "point", shape = 20, size = 5, color = "red", fill = "red")+
  geom_sina(alpha = 0.5)+
  geom_line(data = df_mean, mapping = aes(x = type_detailed, y = average, group = 1))+
  theme(legend.position = "none")+
  xlab("Type Detailed")+
  ylab("Guaranteed Minimum")+
    labs(
    title = "Violin-Sina plot of guaranteed minimum by ICE Detention Facility Type",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
    )
```

# Ghost Beds {#ghost-beds}

Of the 50 facilities that have guaranteed minimums, all but one report a facility Total ADP that is lower than the facility guaranteed minimum (Table \@ref(tab:ghost-bed-by-facility)). Only the Calhoun County Correctional Center has a total ADP that is higher than the facility guaranteed minimum number of beds. Unoccupied guaranteed minimum beds are termed "ghost beds." There are a total of 22960 ghost beds that are located in CDF's, IGSA's, and SPC's. Given this, during FY21 up to April, 67% of Guaranteed Minimum beds are Ghost Beds that are being paid for and are empty.

```{r ghost-bed-by-facility, layout = "l-body-outset"}
# Estimating Ghost Beds
df_ghost_beds <- df %>%
  filter(guaranteed_minimum >0) %>%
  mutate(estimated_ghost_beds = guaranteed_minimum - total_detained) %>%
  arrange(desc(estimated_ghost_beds))

# Ghost Beds by Facility
df_ghost_beds %>% 
    select(c(name,type_detailed, estimated_ghost_beds, guaranteed_minimum, total_detained))%>%
  kable(
    col.names = c("Facility",
                  "Type Detailed",
                  "Estimated Ghost Beds",
                  "Guaranteed Minimums",
                  "Total Detained"
                  ),
    caption="Estimated number of ghost beds by ICE detention facility."
    )%>% 
  scroll_box(height = "300px")
```

Does Total Detained explain the presence and amount of Ghost Beds? The Pearson's product-moment correlation between Ghost Beds and Total Detained is positive, statistically significant, and very large ($r = 0.43, t(48) = 3.30, p < .01$). Using a linear model to predict Ghost Beds by Total Detained (Figure \@ref(fig:scatterplot-ghost-bed-by-total-detained)) explains a statistically significant and moderate proportion of variance ($R^2 = 0.18, F(1, 48) = 10.88, p = 0.002, adj. R^2 = 0.17$). The effect of Total Detained on Ghost Beds is statistically significant and positive ($b = 1.07,t(48) = 3.30, p < .01$). Total Detained explains about 18% of Ghost Beds. This is probably related to the fact that Total Detained predicts the amount of Guaranteed Minimum (Section 3).

```{r cor-gb-td, include=FALSE}
cor.test(df_ghost_beds$estimated_ghost_beds, df_ghost_beds$total_detained, method = "pearson") %>%
  report()

```

```{r lm-gp-td, include=FALSE}
lm(estimated_ghost_beds~total_detained, df_ghost_beds) %>%
  report()
```

```{r scatterplot-ghost-bed-by-total-detained, fig.cap="Scatterplot of Ghost Beds by Total Detained."}
df_ghost_beds %>% 
  ggplot(aes(x=total_detained, y = estimated_ghost_beds)) + 
  geom_point(aes(color = type_detailed )) + 
  geom_smooth(method="lm")+
    labs(title="Scatterplot of Ghost Beds by Total Detained with regression line and 95% \nconfidence interval.",
         caption = "Data source: https://www.ice.gov/detain/detention-management/")
```

Does the amount of Guaranteed Minimum explain the presence and amount of Ghost Beds? The Pearson's product-moment correlation between Ghost Beds and Guaranteed Minimum is positive, statistically significant, and very large ($r = 0.96, t(48) = 22.31, p < .001$). Using a linear model to predict Ghost Beds by Guaranteed Minimum (Figure \@ref(fig:scatterplot-ghost-beds-by-guaranteed-min)) explains a statistically significant and very large proportion of the variance ($R^2 = 0.91, F(1, 48) = 497.86, p < .001, adj. R^2 = 0.91$). The effect of Guaranteed Minimum on Ghost Beds is statistically significant and positive ($b = 0.78, t(48) = 22.31, p < .001$). The presence of Guaranteed Minimums explains 91% of Ghost Beds.

```{r cor-gb-gm, include=FALSE}
cor.test(df_ghost_beds$estimated_ghost_beds, df_ghost_beds$guaranteed_minimum, method = "pearson") %>% 
  report()
```

```{r lm-gb-gm, include=FALSE}
lm(estimated_ghost_beds~guaranteed_minimum, df_ghost_beds) %>% 
  report()
```

```{r scatterplot-ghost-beds-by-guaranteed-min, fig.cap="Scatterplot of Ghost Beds by Guaranteed Minimum"}
df_ghost_beds %>% 
  ggplot(aes(x=guaranteed_minimum, y = estimated_ghost_beds)) + 
  geom_point(aes(color = type_detailed )) + 
  geom_smooth(method="lm")+
      labs(title="Scatterplot of Ghost Beds by Guaranteed Minimum with regression \nline and 95% nconfidence interval.",
         caption = "Data source: https://www.ice.gov/detain/detention-management/")
```

## Ghost Beds by Facility Type {#ghost-beds-by-facility-type}

```{r ghost-bed-sum}
# Sum of Ghost Beds
ghost_bed_sum <- sum(df_ghost_beds$estimated_ghost_beds)
```

Table \@ref(tab:ghost-bed-by-facility-type) provides some summary figures for ghost beds by facility type. Just as all SPC's have guaranteed minimums, all SPC's also have ghost beds, and they make up 10% of the facilities with ghost beds. There are a total of 1293 ghost beds in SPC's representing 5.63% of the total number of ghost beds. SPC's average 258.6 ghost beds per facility which is the lowest among the three types of facilities that have ghost beds.

There are 29 IGSA's that have ghost beds representing 58% of the facilities with Ghost Beds. There are 14,812 Ghost Beds in IGSA facilities making up 64.51% of all ICE Ghost Beds. With an average of 510.76 Ghost Beds per facility, IGSA's have the highest mean Ghost Beds of any of the three types of facilities.

There are 15 CDF's that have Ghost Beds representing 30% of the facilities that have Ghost Beds. There are a total of 6,894 Ghost Beds in CDF's representing 30.03% of all ICE Ghost Beds. Having an average of 459.6 ghost bed per facility, CDF's fall in between IGSA's and SPC's.

```{r ghost-bed-by-facility-type, layout = "l-body-outset"}
# Filter facility types to just those with guaranteed mins.
df_types_ghost <- 
df_types %>% 
  filter(type_detailed == "CDF" | type_detailed == "SPC" | type_detailed == "IGSA") %>% 
  select(c(type_detailed, n))


# Get counts and percent totals of facilities with guaranteed mins.
df_ghost <- df_ghost_beds %>% 
  filter(estimated_ghost_beds > 0) %>% 
  count(type_detailed) %>%
  rename(n_facilities_w_ghost_beds = n) %>% 
 mutate(percent_facilities_by_type_w_ghost_beds = (n_facilities_w_ghost_beds/df_types_ghost$n)*100)%>% 
  mutate(percent_facilities_by_type_w_ghost_beds = round(percent_facilities_by_type_w_ghost_beds, digits = 0)) %>% 
  mutate(percent_facilities_w_ghost_beds = (n_facilities_w_ghost_beds/50)*100)

# Get summary stats of guaranteed mins by facility type.
df_ghost_summaries <- df_ghost_beds %>% 
  filter(estimated_ghost_beds > 0) %>% 
  group_by(type_detailed) %>% 
summarize(total_ghost_beds = sum(estimated_ghost_beds),
            mean_ghost_beds = mean(estimated_ghost_beds),
            sd_ghost_beds= sd(estimated_ghost_beds)) %>% 
   mutate(percent_total_ghost_beds = (total_ghost_beds/ghost_bed_sum)*100, .after = total_ghost_beds) %>% 
  mutate(percent_total_ghost_beds = round(percent_total_ghost_beds, digits = 2))

# Join the three tables
left_join(df_ghost, df_ghost_summaries) %>% 
  kable(
    col.names = c(
      "Type Detailed",
      "n Facilities w. Ghost Beds",
      "% Facilities by Type w. Ghost Beds",
      "% Faciliites w. Ghost Beds",
      "Total Ghost Beds",
      "% Total Ghost Beds",
      "Mean Ghost Beds",
      "SD Ghost Beds"
    ),
    caption = "Ghost beds by facility type."
    )

```

Figure \@ref(fig:ghost-bed-by-facility-type-figure) provides a violin-sina plot of Ghost Beds by facility type and shows that IGSA's have the highest mean number of Ghost Beds and several IGSA facilities have a very large number of Ghost Beds. Figure \@ref(fig:map-ghost-beds) provides a map of Ghost Beds in the continental US. The map shows that Ghost Beds are concentrated on the southern border.

```{r ghost-bed-by-facility-type-figure, fig.cap="Violin-Sina plot of ghost beds by ICE detention facility type."}
df_mean <- df_ghost_beds %>% 
  group_by(type_detailed) %>% 
  summarize(average = mean(estimated_ghost_beds)) %>% 
  ungroup()

df_ghost_beds %>% 
  ggplot(data = ., aes(x= type_detailed, y = estimated_ghost_beds, fill=type_detailed))+
  geom_violin()+
  stat_summary(fun = mean, geom = "point", shape = 20, size = 5, color = "red", fill = "red")+
  geom_sina(alpha = 0.5)+
  geom_line(data = df_mean, mapping = aes(x = type_detailed, y = average, group = 1))+
  theme(legend.position = "none")+
  xlab("Type Detailed")+
  ylab("Estimated Ghost Beds")+
    labs(
    title = "Violin-Sina plot of ghost beds by ICE Detention Facility Type",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
    )
```

```{r map-ghost-beds, fig.cap="Map of estimated number of ghost beds in ICE detention facilities.", preview=TRUE}
# Map of Ghost Beds
ggplot()+
  geom_sf(data=states)+
  geom_point(data=df_ghost_beds, aes(x=longitude, y=latitude, size=estimated_ghost_beds, color = type_detailed))+
  coord_sf(xlim = c(-125,-64), ylim = c(16.5, 49.5), expand = FALSE)+
    labs(
    title = "ICE Detention Facilities",
    subtitle = "Showing Estimated Ghost Beds FY21",
    caption = "Data source: https://www.ice.gov/detain/detention-management/",
    size = "Estimated\nGhost Beds")
```

## Ghost Bed Cost {#ghost-bed-cost}

As of April, the ICE's FY21 year to date data table lists 51 facilities with Guaranteed Minimums but unfortunately does not supply the per diem rate for any of those contracts. The DHS budget request for FY21 provides information on 27 facilities with guaranteed minimums [@dhs2020, ICE-O&S-137], and among those facilities (n=27) the mean per diem is \$123.04 (SD=43.02). Presumably, the Guaranteed Minimums at the remaining 23 facilities were established after the FY21 budget request. For the purpose of estimating the total cost of Guaranteed Minimums and unused Ghost Beds, for the 23 facilities for which Guaranteed Minimums are unlisted the average of the Guaranteed Minimums per diem rate for the 27 facilities listed in the FY21 budget request are used [@dhs2020, ICE-O&S-137].

```{r}
guaranteed_min_per_diem <- read_csv("https://raw.githubusercontent.com/n8craig/ICE-Detention-Year-End-Reports/main/data/ICE_FY21_budget_guaranteed_minimums.csv")
```

```{r}
ghost_beds_join1 <- df_ghost_beds %>% 
  select(c(name, total_detained,
           type_detailed,
           guaranteed_minimum,
           estimated_ghost_beds))



ghost_beds_join2 <- guaranteed_min_per_diem %>%
  janitor::clean_names() %>% 
  select(c(name, guaranteed_minimum_per_diem))

df_ghost_bed_joined <- left_join(ghost_beds_join1, ghost_beds_join2)
```

```{r}
df_ghost_bed_joined <- df_ghost_bed_joined %>% 
  mutate(guaranteed_min_cost = guaranteed_minimum * guaranteed_minimum_per_diem) %>% 
  mutate(ghost_bed_cost = estimated_ghost_beds* guaranteed_minimum_per_diem)
```

```{r}
 
  guaranteed_min_cost <- sum(df_ghost_bed_joined$guaranteed_min_cost)
  
  ghost_bed_cost <- sum(df_ghost_bed_joined$ghost_bed_cost)
```

Based on reported Guaranteed Minimum per diem rates, and average rates where not reported, the total daily cost of Guaranteed Minimum ICE detention beds is \$`r guaranteed_min_cost` and the total estimated cost of Ghost Beds is \$`r ghost_bed_cost`.

Figure \@ref(fig:guaranteed-min-per-diem) shows the Guaranteed Minimum per diem rate by facility type. SPC's tend to have higher Guaranteed Minimum per diem costs than either CDF's or IGSA's. One reason that a large number of IGSA's guaranteed minimums hover around a per diem of 125 is that many of these facilities are not listed in the FY21 year to date figures and the values are based on averages. *It would be good to determine which kinds of facilities have new Guaranteed Minimums*. Figure \@ref(fig:guaranteed-minimum-average-cost-by-facility-type) shows the average Guaranteed Minimum cost by facility type. On average, most facilities have roughly similar costs even though IGSA's and CFS's are larger. The leveling of average cost may be related to the higher per diems for SPC's. Figure \@ref(fig:ghost-bed-cost-by-facility-type) shows the average Ghost Bed cost by facility type. SPC's have slightly lower costs than IGSA's and CDF's.

```{r guaranteed-min-per-diem, fig.cap="Violin-sina plot of Guaranteed Minimum per diem rate by ICE detention Type Detailed."}
df_mean <- df_ghost_bed_joined %>% 
  group_by(type_detailed) %>% 
  summarize(average = mean(guaranteed_minimum_per_diem)) %>% 
  ungroup()


df_ghost_bed_joined %>% 
  ggplot(data = ., aes(x= type_detailed, y = guaranteed_minimum_per_diem, fill=type_detailed))+
  geom_violin()+
  stat_summary(fun = mean, geom = "point", shape = 20, size = 5, color = "red", fill = "red")+
  geom_sina(alpha = 0.5)+
  geom_line(data = df_mean, mapping = aes(x = type_detailed, y = average, group = 1))+
  theme(legend.position = "none")+
  xlab("Type Detailed")+
  ylab("Guaranteed Minimum Per Diem Cost")+
    labs(
    title = "Violin-Sina plot of Guaranteed Minimum Per Diem Rate by ICE Detention \nFacility Type",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
    )
```

```{r guaranteed-minimum-average-cost-by-facility-type, fig.cap="Violin-sina plot of average Guaranteed Minimum Cost by ICE detention Type Detailed."}
df_mean <- df_ghost_bed_joined %>%
  group_by(type_detailed) %>% 
  summarize(average = mean(guaranteed_min_cost)) %>% 
  ungroup()

df_ghost_bed_joined %>% 
  ggplot(data = ., aes(x= type_detailed, y = guaranteed_min_cost, fill=type_detailed))+
  geom_violin()+
  stat_summary(fun = mean, geom = "point", shape = 20, size = 5, color = "red", fill = "red")+
  geom_sina(alpha = 0.5)+
  geom_line(data = df_mean, mapping = aes(x = type_detailed, y = average, group = 1))+
  theme(legend.position = "none")+
  xlab("Type Detailed")+
  ylab("Guaranteed Minimum Cost")+
    labs(
    title = "Violin-Sina plot of Guaranteed Minimum Cost by ICE Detention \nFacility Type",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
    )
```

```{r ghost-bed-cost-by-facility-type, fig.cap="Violin-sina plot of Ghost Bed Cost by ICE detention Type Detailed."}
df_mean <- df_ghost_bed_joined %>%
  group_by(type_detailed) %>% 
  summarize(average = mean(ghost_bed_cost)) %>% 
  ungroup()

df_ghost_bed_joined %>% 
  ggplot(data = ., aes(x= type_detailed, y = ghost_bed_cost, fill=type_detailed))+
  geom_violin()+
  stat_summary(fun = mean, geom = "point", shape = 20, size = 5, color = "red", fill = "red")+
  geom_sina(alpha = 0.5)+
  geom_line(data = df_mean, mapping = aes(x = type_detailed, y = average, group = 1))+
  theme(legend.position = "none")+
  xlab("Type Detailed")+
  ylab("Ghost Bed Cost")+
    labs(
    title = "Violin-Sina plot of Ghost Bed Cost by ICE Detention \nFacility Type",
    caption = "Data source: https://www.ice.gov/detain/detention-management/"
    )
```
