---
title: "ICE Detention COVID-19 Maps"
description: |
  Animated visualizations of COVID-19 infections in ICE detention over time.
author:
  - name: Nathan Craig
    url: https://ncraig.netlify.app/
date: 05-23-2021
categories:
  - immigration
output:
  distill::distill_article:
    self_contained: false
    pandoc_args: ["--number-sections"]
    df_print: paged
    toc: true
    number_sections: false
bibliography: packages.bib
link-citations: yes
repository_url: https://github.com/n8craig/ncraig
creative_commons: CC BY-NC
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
# Data loading
library(readr)

# Data wrangling
library(tidyverse)
library(janitor)

# Spatial data
library(sf)
library(maps)

# Plotting
library(ggplot2)
library(gganimate)
library(magick)
library(patchwork)
```

```{r}
# Load facility specific data
df <- read_csv("https://raw.githubusercontent.com/n8craig/ICE-COVID-19/main/data/covid_by_facility.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y"), 
        `Confirmed cases currently under isolation or monitoring` = col_number(), 
        `Detainee deaths` = col_number(), 
        `Total confirmed COVID-19 cases` = col_number()))
df <- clean_names(df)

locations <- read_csv("https://raw.githubusercontent.com/n8craig/ICE-COVID-19/main/data/facility_management_location.csv")
locations <- clean_names(locations)

df <- left_join(df, locations)
df <- df %>% mutate(manager = as.factor(manager))


# Load summary data
df_summary <- read_csv("https://raw.githubusercontent.com/n8craig/ICE-COVID-19/main/data/covid_summaries.csv",
                       col_types = cols(Date = col_date(format = "%m/%d/%Y")))
df_summary <- clean_names(df_summary)
df_summary$date <- as.Date(df_summary$date)
```

# COVID-19 Cases in ICE Detention Over Time

The following maps and graph are generated with data scraped from the [ICE Guidance on COVID-19](https://www.ice.gov/coronavirus) website. The scraped data are maintained in a GitHub [repository](https://github.com/n8craig/ICE-COVID-19). COVID-19 data reported by ICE are linked with geocoded facility locations. COVID-19 data are assembled in "long" format, plotted with `ggplot` [@R-ggplot2], and animated with the `gganimate` [@R-gganimate] library.

## Cumulative COVID-19 Map

A map of cumulative COVID-19 cases by ICE detention facility over time.

<aside>

```{r}
df <- df %>% 
  filter(type !="Hotel")

fig <- 
  df %>% 
  drop_na(type) %>% # Long term should resolve these records that don't match.!
  ggplot(aes(x=longitude, y=latitude))+
  borders("state")+
  geom_point(alpha = 0.2,
             aes(color = type,
                 size = total_confirmed_covid_19_cases,
                 frame = date,
                 group = 1L
                 ))+
  guides(color = guide_legend(override.aes = list(alpha= 1)))+
  coord_sf(xlim = c(-124,-68), ylim = c(24, 49.5), expand = TRUE)+
  theme(legend.position = "right")+
  labs(
    title = "Cumulative COVID-19 Cases: {frame_time}",
    caption = "Data source: https://www.ice.gov/coronavirus"
    )+
  guides(size = FALSE, alpha = FALSE)
fig
```

</aside>

```{r}

df <- df %>% 
  filter(type !="Hotel")

fig <- 
  df %>% 
  drop_na(type) %>% # Long term should resolve these records that don't match.!
  ggplot(aes(x=longitude, y=latitude))+
  borders("state")+
  geom_point(alpha = 0.2,
             aes(color = type,
                 size = total_confirmed_covid_19_cases,
                 frame = date,
                 group = 1L
                 ))+
  guides(color = guide_legend(override.aes = list(alpha= 1)))+
  coord_sf(xlim = c(-124,-68), ylim = c(24, 49.5), expand = TRUE)+
  transition_time(date)+
  ease_aes()+
  theme(legend.position = "right")+
  labs(
    title = "Cumulative COVID-19 Cases: {frame_time}",
    caption = "Data source: https://www.ice.gov/coronavirus"
    )+
  guides(size = FALSE, alpha = FALSE)

animate(fig, fps=4, end_pause = 7)
```

## Cumulative COVID-19 Line Plot

```{r}
fig <- 
df_summary %>% 
  ggplot(aes(x=date, y=total_cumulative_covid_19))+
  geom_line()+
  geom_point()+
  transition_time(date)+
  transition_reveal(date)+
  labs(
    title = "Cumulative COVID-19 Cases: {frame_along}",
    caption = "Data source: https://www.ice.gov/coronavirus"
  )+
  xlab("Date")+
  ylab("Cumulative COVID-19 Cases in ICE Detention")

animate(fig, fps=4, end_pause = 7)
```

## Total Detained and Active Cases Line Plot Over Time

The following is a pair of line plots showing detained population and active COVID-19 cases in custody over time. Comparing the plots one can see that between July and January 2020 as the detained population decreased the number of active COVID-19 cases in custody declined. Likewise, one can see that in late March 2021 as the total detained population began to rise the number of active COVID-19 cases in custody also rose and that by April 2021 both the detained population and active COVID-19 cases in ICE detention are rising rapidly.

```{r, layout = "l-page"}
# ICE Detention Population Over Time
total_detained <- ggplot(df_summary, aes(x=date, y=total_detained))+
  geom_line()+
  geom_point()+
  geom_smooth()+
  labs(
    title = "ICE Detention Population \nOver Time",
    subtitle = "Smoothed trend line added"
  )+
  xlab("Date")+
  ylab("Total Detained Population")


# Total Confirmed COVID-19 Over Time
active_in_custody <- ggplot(df_summary, aes(x=date, y=total_covid_19_confirmed_in_custody))+
  geom_line()+
  geom_smooth()+
   labs(
    title = "Confirmed COVID-19 Cases \nin ICE Detention Over Time",
    subtitle = "Smoothed trend line added",
    caption = "Data source: https://www.ice.gov/coronavirus"
  )+
  xlab("Date")+
  ylab("Total Active COVID-19 Cases in Custody")

total_detained + active_in_custody
```

## Active COVID-19 Map

A map of active COVID-19 cases by ICE detention facility over time.

```{r}
df <- df %>% 
  filter(type !="Hotel")

fig <- 
  df %>% 
  drop_na(type) %>% # Long term should resolve these records that don't match.!
  ggplot(aes(x=longitude, y=latitude))+
  borders("state")+
  geom_point(alpha = 0.2,
             aes(color = type,
                 size = confirmed_cases_currently_under_isolation_or_monitoring,
                 frame = date,
                 group = 1L
                 ))+
  guides(color = guide_legend(override.aes = list(alpha= 1)))+
  coord_sf(xlim = c(-124,-68), ylim = c(24, 49.5), expand = TRUE)+
  transition_time(date)+
  ease_aes()+
  theme(legend.position = "right")+
  labs(
    title = "Active COVID-19 Cases: {frame_time}",
    caption = "Data source: https://www.ice.gov/coronavirus"
    )+
  guides(size = FALSE, alpha = FALSE)

animate(fig, fps=4, end_pause = 7)
```

```{r}
knitr::write_bib(file = "packages.bib")
```
